Beyond the coastal periphery: chert use and procurement patterns in the Upper Paleolithic of Southwesternmost Portugal

### 2024-01-08 analysis/paper/paper.qmd
First draft

---
title: "Title Goes Here"
author:
  - Jane Doe:
      correspondence: "yes"
      email: janedoe@fosg.org
      orcid: 0000-0003-1689-0557
      institute:
        - fosg
        - fop
  - John Q. Doe:
      institute: fosg
      orcid: 0000-0003-1689-0558
  - Peder Ås:
      institute: fosg
      orcid: 0000-0003-1689-0559
  - Juan Pérez:
      orcid: 0000-0003-1689-0551
      institute:
        - name: Acme Corporation
  - Max Mustermann:
      orcid: 0000-0003-1689-0552
institute:
  - fosg:
      name: Formatting Open Science Group
      address: 23 Science Street, Eureka, Mississippi, USA
  - fop: Federation of Planets
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
number-sections: true
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
---

@@

Knappable raw materials are a key element used to understand the lifeways of past hunter-gatherers throughout time, as lithics are ubiquitous throughout prehistory. More than mere stones with which stone tools are produced, these resources are intimately connected to the group’s technological, social, and cultural organization, possibly with an impact on the group's survivability [@oestmoFormalModelingApproach2017]. Several key topics to understand hunter-gatherer behavior and organization have been explored through raw material analysis, such as modalities of procurement and mobility strategies [@ambrose_social_1990; @binford_organization_1979; @binford_righteous_1985; @gould_lithic_1985; @gould_lithic_1985; @kuhn_unpacking_1991; @mccall_behavioral_2007], occupation types and their duration [@kuhn_upper_2004; @surovell_toward_2009]) to the establishment and dimension of social networks [@whallon_social_2006] and exchange between groups or individuals [@wurzHowiesonsPoortBacked1999; @gamblePalaeolithicSocietiesEurope1999]. This has frequently been done through the systematic characterization of raw materials (both geologically available and archaeological), identification of corresponding sources for a clear differentiation between local and long-distance raw materials, and the study of their proportions within a site, often throughout different occupations and chronologies. These approaches and concepts have been extensively applied to the study of lithic assemblages from prehistoric archaeological sites in the Iberian Peninsula (++). Frequently coupled with technological analyses, the study of raw materials, especially chert, through a wide range of methods, such as macroscopic, petrographic, and geochemical, have recurrently aided in answering key questions to Iberian archaeology, such as the disappearance of Neanderthals and consequent replacement by Modern Humans [@aubry_far_2022]. Raw material studies have also contributed to understanding how different groups during different chronological periods organized their technology, managed time, and risk (through the embedded/direct procurement approach), and chose resources [@matias2016; @matias_o_2012; @pereira_raw_2021; @pereiraLithicEconomyTerritory2016]. Furthermore, these studies frequently highlight long-distance procurement even if in low proportions at times. In fact, raw material studies and the distinction between local/regional and exotic have also been successfully applied to identify long-distance contacts (\>150 km) during the Upper Paleolithic in the Côa Valley which has been interpreted as the result of reciprocal exchange [@aubry_raw_2004-1; @aubry_upper_2016]. Although commonly known and identified during late prehistory [@nocete_circulation_2005; @rodriguezPrehistoricFlintExploitations2011; @costa_lithic_2022], the identification of these contacts has expanded our understanding of often large-scale social networks and contacts during the Upper Paleolithic in central Portugal, to paint a picture previously outlined of a wide interconnected territory. However, unlike other regions of the Iberian Peninsula such as central Portugal, and despite the potential of raw material studies to provide key data on the lifeways and organization of Paleolithic hunter-gatherers through time, as well as the mobility range and existence of social networks, the behaviors of management and procurement of raw materials are still in a preliminary phase in southern Portugal, especially in the Algarve region. Although much is known about the technological organization of Paleolithic hunter-gatherer groups in southwestern Portugal, especially due to the archaeological site of Vale Boi (see ++), raw material studies have been scarcer. Previous studies applied macroscopic methods to characterize the cherts from Vale Boi. These preliminary results have shown the presence of a large component of local cherts from the Jurassic limestone formations in conjunction with a less abundant and variable allochthonous chert component throughout the several Upper Paleolithic occupations of the site [@bicho_ecodynamics_2013; @pereira_territory_2016-1; @verissimo_jazidas_2004]. @pereira_territory_2016-1 suggest that the exogenous group may have been the result of embedded procurement within a 16 km radius, while the allochthonous group may have been the result of transportation during trips to other areas of the Algarve, Alentejo, or even south Spain. The presence of chert from the Cretaceous formations of central Portugal has also been suggested [@bicho_ecodynamics_2013]. Furthermore, previous studies aiming to characterize the local and regional chert sources in the Algarve region have established a comparative macroscopic and petrographic database [@belmiro2023; @pereira_presenting_2016-1], a key element to raw material studies, especially to distinguish between local and non-local raw materials and understanding changes in the archaeological record [@pop_simulating_2015; @surovell_toward_2009]. These preliminary results and the existence of regional reference collections show the potential for a systematic raw material study at the site, using different methods and approaches, applied to the lithic assemblages throughout the several Upper Paleolithic occupations, to understand how these hunter-gatherer groups behaved through time: how they explored their landscape and lithic resources, where they moved across a territory nestled between water and land, and with whom did they establish social networks. To answer these questions, we aim to characterize the raw material procurement and mobility patterns of hunter-gatherer groups throughout the Upper Paleolithic of southern Portugal. By studying the chalcedony and chert artifacts of the UP levels of Vale Boi we aim to characterize these raw materials macroscopically and petrographically, to identify different chert groups and their respective sources.

@@

Vale Boi is located on the western coast of the Algarve region (south Portugal), in a small valley that connects south to the Atlantic current coast at around 2 km and is bordered by a limestone hill. This hill is characterized by limestone exposures that form rocks shelters facing west and southwest [@cascalheira_tecnologia_2010]. The site extends for more than 10 000 sq m through the slope of the valley, through which three main occupation areas have been identified: Slope, Shelter and Terrace. The excavation methodology included the recording of all artifacts with dimensions superior to 2 cm or complete small artifacts (i.e., bladelets) with a total station, and was applied to all main areas. Several occupations have been identified in both the Terrace and the Shelter area. The Terrace includes occupations from the Holocene to the Upper Paleolithic, resulting in the identification of eight main lithostratigraphic units. The UP occupations start with the Gravettian (levels 8-6), followed by Proto-solutrean (bottom of 5 and 5/4E, sensu Belmiro et al., 2021), and Solutrean occupations (4D-4 and 3B). The Shelter area shows four main lithostratigraphic units, with Magdalenian, Solutrean and Gravettian occupations. From these, the Solutrean levels show the largest occupation, with three, well-preserved archaeological horizons [@cascalheira_vale_2013].

@@

The Algarve region is a complex territory, characterized by several geomorphic sub-regions and geological units, which provide different lithological resources used throughout prehistory. There are two main geological units, the South Portuguese Zone (SPZ) in the north sector of Algarve and the Algarve basin to the south. The SPZ is characterized by schists, greywacke and quartzites, while the Algarve basin is characterized by limestones, dolomites, and marls. It is within these limestones and dolomitized limestones from the Triassic to the Upper Jurassic where chalcedony and chert nodules and beds can be found. Chalcedony is located in a 15 km radius from Vale Boi, in the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes [@bicho_ecodynamics_2013]. The cherts can be found in different Jurassic formations throughout the Algarve, but are most frequently known for the Lower Jurassic beds and small nodules in the limestone cliffs at ++ km distance from the site, which have been previously studied and characterized [@ribeiro_evolucao_2005; @belmiro2023]. Chert is also available in the less known Upper Jurassic limestone formation in western Algarve (++ km from the site), and in the center and eastern portion of the Algarve, in Middle and Upper Jurassic formations with chert nodules [@belmiro2023].

@@

To better characterize the different chert groups identified in the macroscopic analysis, 22 thin sections were prepared at Thin Section Lab (Toul, France) and analyzed at the ICArEHB laboratory. The samples chosen for this analysis include at least one sample from each group present throughout the stratigraphy of the Terrace area, from different archaeological levels (@tbl-thinsections). For groups that showed macroscopic variability, several thin sections were analyzed to ensure the collection of petrographic data that reflected this variability. Petrographic thin sections were analyzed by polarized light microscopy (Nikon LV100ND) to determine the mineral composition, textural characteristics, allochems, and bioclasts, as well as any alterations.

| Type               | Sample ID | Level | TS \# |
|--------------------|-----------|-------|-------|
| Chalcedony         | I19-1615  | 4D    | 8     |
| Type 2             | I19-2835  | 5     | 17    |
| Type 2 (rmu_17)    | I19-2226  | 4E    | 18    |
| Type 2 (altered)   | H18-1938  | 4D    | 20    |
| Type 2 (reddening) | H19-2426  | 4D    | 3     |
| Type 2B            | H21-3095  | 4C    | 6     |
| Type 2B            | H21-4234  | 5     | 7     |
| Type 15            | I20-3160  | 5     | 10    |
| Type 7             | H20-2441  | 4E    | 19    |
| Type 12            | I21-3252  | 4E    | 14    |
| Type 12B           | H18-2708  | 5     | 2     |
| CPT                | I19-3350  | 6     | 9     |
| CPT                | J18-778   | 5     | 13    |
| CPT                | I20-3689  | 6     | 16    |
| Sub-type LLB       | I20-3951  | 6     | 11    |
| Sub-type 11E       | H19-4074  | 6     | 4     |
| Type 11F           | H19-2924  | 4E    | 1     |
| Type 13 (01b; 01a) | L19-64    | 5     | 12    |
| Type 13 (04)       | J18-1264  | 8     | 22    |
| Type 13B           | I21-2966  | 4C    | 15    |
| Oolitic            | H19-4216  | 6     | 5     |
| White patinated    | H20-4166  | 6     | 21    |

: A table with thin sections {#tbl-thinsections}

@@

T_dataset <- T_dataset %>% 
  #filter(Level %in% c("4", "4C", "4D", "4E", "5", "6", "7"))
  filter(Level !="4B") %>%  # Filtering out levels with barely any artifacts
  mutate(Level = case_when(Level == "7C" ~ "7", # Collapsing samples from Level 7C into 7
                           TRUE ~ Level)) %>% 
  mutate(Level = case_when(Level == 8 ~ "7", # Collapsing two samples from Level 8 into 7, since 8 disappears in field_data
                           TRUE ~ Level))

@@

S_RM_data <- S_RM_data %>%
  mutate_at("ID", str_replace, "VB", "") %>% # Remove "VB" from ID
  mutate(RMTYPE = case_when(
    RMTYPE == "Type 2" ~ "TYPE 2",
    .default = RMTYPE))

@@

T_table <- phase_dataset %>%
  filter(RMTYPE !="INDET") %>% 
  mutate(RMTYPE = factor(RMTYPE, levels = c("Chalcedony","TYPE 2", "TYPE 15", "TYPE 7",
                                        "TYPE 12", "TYPE 13","TYPE 13B", 
                                        "TYPE 11", "TYPE 11F", "Oolitic", "GB", "TL"))) %>%
  tabyl(RMTYPE, Level) %>%
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(2) %>%
  adorn_ns("front")

@@

The three main used raw materials at the archaeological site of Vale Boi are quartz, greywacke, and chert. Other raw materials such as chalcedony, dolerite and schists have also been identified. Using the field database with all individually coordinated lithic artifacts, we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials. @fig-rm-comparison shows that for the selected levels, chert represents only \~20-15% of the lithic assemblage. Based on previous lithic studies in the terrace area, the remaining 80% of other raw materials are mostly composed of quartz and greywacke \[++\]. Although no systematic raw material research has been applied to the latter raw materials, they have been interpreted as local and readily available in the proximity of the site \[++\]. Frequently, the quartz and greywacke assemblages include large amounts of shattered and/or unknapped chunks or slabs. Previous studies from the terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz [@belmiro_gravettian-solutrean_2021]. Based on this, despite the low amounts of individually plotted chert artifacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits. The importance of chalcedony has also been shown in its use to produce formal tool kits during specific technocomplexes at the site [@belmiro_gravettian-solutrean_2021].

@@

context_short <- context %>% 
  filter(Level %in% c("4", "4C", "4D", "4E", "5", "6", "7"))
dataset_short <- T_dataset %>% 
  select(ID, RMTYPE)
T_short <- full_join(context_short, dataset_short, by = "ID")
T_short["RMTYPE"][is.na(T_short["RMTYPE"])] <- "Other"

T_short <- T_short %>% 
 mutate(Rawmaterial = case_when(
    RMTYPE == "Other" ~ "Other",
    RMTYPE == "Chalcedony" ~ "Chalcedony",
    RMTYPE != "Other" | RMTYPE != "Chalcedony" ~ "Chert"))

@@

Through the macroscopic and petrographic analysis, we identified 10 main chert/chalcedony groups, present throughout the stratigraphy in the Terrace and Shelter area. A single different group was identified in the Shelter area, with limited presence in the Terrace (n=2), and 14 varieties of chert and 2 chalcedony composed of one to three samples were identified in several levels of the Terrace and Shelter. The latter varieties were collapsed into a category named TL (Trace lithotypes). Through the comparison with the reference collections, we identified cherts congruent with the local sources and cherts which are not local.

The raw materials which are congruent with local sources are the chalcedony and massive, micro-cryptocrystalline cherts. The chalcedony is a massive, fibrous chalcedony without allochems. It is characterized by a colorless and translucent appearance, often partially or completely white patinated. It is characterized by frequent large crystallizations, irregularities and fractures which impact the knapping. It is macroscopically and petrographically congruent with the hydrothermal chalcedony found in lenses in Triassic formations, also frequently characterized by irregularities and fractures.

The cherts are massive micro-cryptocrystalline quartzes from a marine environment. Macroscopically, these groups show variability which resulted in their separation in four different macroscopic groups (Group 2 to 5). In general, these are opaque to sub-translucent cherts, with colors varying brown, yellow, pale red and grey, with the frequent presence of spicules and iron oxides. The differences are in their colors, nodule morphology, frequency, and variety of bioclasts, something already identified in the geological samples [@belmiro2023]. Petrographically, however, these cherts are homogeneous, except for a small number of artifacts with a banded variety within group 5. The macroscopic separation was maintained for its possible usefulness in identifying preferred types of nodules or macroscopic varieties through time and for future correlation with technological data. These cherts are characterized by wackestone textures, with accessory amounts of macrocrystalline quartz and fibrous chalcedony, frequently found replacing fossils and occasionally dolomite. Group 12 shows higher percentages of micrite/sparite. In all samples, fossils range from common to very frequent, albeit frequently unidentifiable. When identifiable in thin-section, they are sponge spicules, with the rare occurrence of possible foraminifera and ostracod in a sample from type 12. Porosity varies between 10% and 1%, of vug type. Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiro2023], we interpret Groups 2, 15, 7 and 12 to be local, as they are congruent with the described characteristics. Groups 2 and 15 are congruent with the frequently available brown/yellow and red nodules and beds, ranging from uncommon to very frequent fossils. The macroscopic similarities of Groups 7 and 12 are more limited, as they show similarities to specific nodules, rarely found in the known outcrops, and mostly located in the Ferrel area. As such, we interpret these cherts as local, but from nodules and outcrops that are no longer frequently available or visible in the landscape. This is especially relevant, since the Ferrel outcrops are mostly destroyed and can be found in poor preservation conditions, and with small, broken nodules in the surface, frequently altered [@belmiro2023].

Five chert types, not congruent with local sources, were identified (Groups 6 to 10), including the TL group.

>>>>

---
title: "Beyond the coastal periphery: chert use and procurement patterns in the Upper Paleolithic of Southwesternmost Portugal"
author:
  - Joana Belmiro:
      correspondence: "yes"
      email: jfbelmiro@ualg.pt
      institute:
        - ICArEHB
  - Xavier Terradas:
      institute: 
        - CSIC
  - Salvador Dominguez-Bella:
      institute: 
        - UCA
  - JOão Cascalheira:
      institute:
        - ICArEHB
title-block-published: "Last updated" 
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
number-sections: true
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
---

@@

Knappable raw materials play a crucial role in understanding the lifeways of past hunter-gatherers across time, given the ubiquitous presence of lithics throughout prehistory. More than mere stones with which stone tools are produced, these resources are intimately connected to a group’s technological, social, and cultural organization, potentially influencing the group's overall survivability [@oestmoFormalModelingApproach2017]. Several key topics to the study of hunter-gatherer behavior and organization have been explored through raw material analysis, such as modalities of procurement and mobility strategies [@ambrose_social_1990; @binford_organization_1979; @binford_righteous_1985; @gould_lithic_1985; @gould_lithic_1985; @kuhn_unpacking_1991; @mccall_behavioral_2007], occupation types and their duration, [@kuhn_upper_2004; @surovell_toward_2009] to the establishment and dimension of social networks [@whallon_social_2006] as well as exchanges between groups or individuals [@wurzHowiesonsPoortBacked1999; @gamblePalaeolithicSocietiesEurope1999]. This has frequently been done through the systematic characterization of geological and archaeological raw materials, the identification of corresponding sources for a clear differentiation between local and long-distance raw materials, and the study of their proportions within a site, often throughout different occupations and chronologies. These approaches and concepts have also been extensively applied to the study of lithic assemblages from prehistoric archaeological sites in the Iberian Peninsula (++). Frequently coupled with technological analyses, the study of raw materials, especially chert, through a wide range of methods such as macroscopic, petrographic, and geochemical, have recurrently aided in answering key questions to Iberian archaeology. One such question is the disappearance of Neanderthals and consequent replacement by Modern Humans [@aubry_far_2022]. Raw material studies have also contributed to understanding how different groups organized their technology, managed time, and risk (through the embedded/direct procurement approach) and chose resources during different chronological periods [@matias2016; @matias_o_2012; @pereira_raw_2021; @pereiraLithicEconomyTerritory2016]. Furthermore, these studies frequently highlight long-distance procurement, even if in low proportions at times. In fact, raw material studies and the distinction between local/regional and exotic have also been successfully applied to identify long-distance contacts (\>150 km) during the Upper Paleolithic (UP) in the Côa Valley, interpreted as the result of reciprocal exchange [@aubry_raw_2004-1; @aubry_upper_2016]. Although commonly known and identified during late prehistory [@nocete_circulation_2005; @rodriguezPrehistoricFlintExploitations2011; @costa_lithic_2022], the identification of these contacts has expanded our understanding of often large-scale social networks and contacts during the UP in central Portugal, painting a picture of a wide interconnected territory. However, unlike other regions of the Iberian Peninsula such as central Portugal, and despite the potential of raw material studies to provide key data on the lifeways and organization of Paleolithic hunter-gatherers through time, as well as the mobility range and existence of social networks, the behaviors of management and procurement of raw materials are still in a preliminary phase in southern Portugal, especially in the Algarve region. Although much is known about the technological organization of Paleolithic hunter-gatherer groups in southwestern Portugal, especially due to the archaeological site of Vale Boi [see @cascalheira_vale_2013; @bicho_ecodynamics_2013; @marreiros2015], raw material studies have been scarce. Previous studies applied macroscopic methods to characterize the cherts from Vale Boi. Preliminary results showed the presence of a large component of local cherts from the Jurassic limestone formations, along with a less abundant and variable allochthonous chert component throughout the several UP occupations of the site [@bicho_ecodynamics_2013; @pereira_territory_2016-1; @verissimo_jazidas_2004]. @pereira_territory_2016-1 suggest that the exogenous group may have been the result of embedded procurement within a 16 km radius, while the allochthonous group may have been the result of transportation during trips to other areas of the Algarve, Alentejo, or even south Spain. The presence of chert from the Cretaceous formations of central Portugal has also been suggested [@bicho_ecodynamics_2013]. Furthermore, previous studies aiming to characterize the local and regional chert sources in the Algarve region have established a comparative macroscopic and petrographic database [@belmiro2023; @pereira_presenting_2016-1], a key element in raw material studies, especially to distinguish between local and non-local raw materials and understanding changes in the archaeological record [@pop_simulating_2015; @surovell_toward_2009]. These preliminary results, coupled with the existence of regional reference collections, show the potential for a systematic raw material study at the site, using different methods and approaches and applied to the lithic assemblages throughout the several Upper Paleolithic occupations, to understand how these hunter-gatherer groups behaved through time: how they explored their landscape and lithic resources, where they moved across a territory nestled between water and land, and with whom did they establish social networks. To address these questions, we aim to characterize the raw material procurement and mobility patterns of hunter-gatherers throughout the Upper Paleolithic of southern Portugal. By studying the chalcedony and chert artifacts of the UP levels of Vale Boi we aim to characterize these raw materials macroscopically and petrographically, to identify different chert groups and their souces.

@@

Vale Boi is located on the western coast of the Algarve region in southern Portugal, within small valley that connects south to the Atlantic current coast at around 2 km. It is bordered by a limestone hill, characterized by limestone exposures that form rocks shelters facing west and southwest [@cascalheira_tecnologia_2010]. The site extends for more than 10,000 sq m along the slope of the valley, through which three main occupation areas have been identified: Slope, Shelter and Terrace. The excavation methodology included the recording of all artifacts with dimensions superior to 2 cm or complete small artifacts (i.e., bladelets) with a total station, and was applied to all main areas. Several occupations have been identified in both the Terrace and the Shelter area. The Terrace includes occupations from the Holocene to the Upper Paleolithic, resulting in the identification of eight main lithostratigraphic units where several UP occupations were identified, from the Gravettian, Proto-Solutrean and Solutrean. The Shelter area shows four main lithostratigraphic units, with Magdalenian, Solutrean and Gravettian occupations. From these, the Solutrean levels show the largest occupation, with three, well-preserved archaeological horizons [@cascalheira_vale_2013].

TABLE: Table with short layer description, attributed technocomplex, and radiocarbon dates. Will need complete field layer description and the dates we currently have for VB from JC.

MAP: With geological context, location of VB, Iberian Peninsula small map.

@@

The Algarve region is a complex territory, marked by several geomorphic sub-regions and geological units that provide different lithological resources used throughout prehistory. Two main geological units characterize the region: the South Portuguese Zone (SPZ) in the north sector of Algarve and the Algarve basin to the south. The SPZ is characterized by schists, greywacke and quartzites, while the Algarve basin features limestones, dolomites, and marls. It is within these limestones and dolomitized limestones from the Triassic to the Upper Jurassic where chalcedony and chert nodules and beds can be found. Chalcedony is located approximately 10 km from Vale Boi, in the marl-carbonated Triassic formations as lenses or nodules and was formed through hydrothermal processes [@bicho_ecodynamics_2013]. The cherts can be found in several Jurassic formations throughout the Algarve, but are frequently known for the Lower Jurassic beds and small nodules in the limestone cliffs at \~20 km from the site, which have been previously studied and characterized [@ribeiro_evolucao_2005; @belmiro2023]. Chert is also available in the less known Upper Jurassic limestone formation in western Algarve (++ km from the site), and in the center and eastern portion of the Algarve, in Middle and Upper Jurassic formations with chert nodules [@belmiro2023].

QUESTION: Should I include a short description of the cherts in the text, or do this through an image or table?

@@

To better characterize the different chert groups identified in the macroscopic analysis, 22 thin sections were prepared at Thin Section Lab (Toul, France) and analyzed at the ICArEHB laboratory. The samples chosen for this analysis include at least one sample from each group present throughout the stratigraphy of the Terrace area, from different archaeological levels (@tbl-TS-list). For groups that showed macroscopic variability, several thin sections were analyzed to ensure the collection of petrographic data that reflected this variability. Petrographic thin sections were analyzed by polarized light microscopy (Nikon LV100ND) to determine the mineral composition, textural characteristics, allochems, and bioclasts, as well as any alterations.

```{r}
#| label: tbl-TS-list
#| tbl-cap: A table with a thin section list.

TS_list <- read_csv(here::here('analysis/data/raw_data/TS_short_list.csv'))

kable(TS_list)

@@

T_dataset <- T_dataset %>% 
  #filter(Level %in% c("4", "4C", "4D", "4E", "5", "6", "7"))
  filter(Level !="4B") %>%  # Filtering out levels with barely any artifacts
  mutate(Level = case_when(Level == "7C" ~ "7",
                           Level == 8 ~ "7", # Collapsing samples from Level 7C into 7 and two samples from Level 8 into 7, since 8 disappears in field_data
                           TRUE ~ Level)) %>% 
  mutate(RMTYPE = case_when(RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
                            RMTYPE == "TYPE 15" ~ "TYPE 3",
                            RMTYPE == "TYPE 7" ~ "TYPE 4",
                            RMTYPE == "TYPE 12" ~ "TYPE 5",
                            RMTYPE == "TYPE 11" ~ "TYPE 6",
                            RMTYPE == "TYPE 13" ~ "TYPE 7",
                            RMTYPE == "TYPE 11F" ~ "TYPE 8",
                            RMTYPE == "TYPE 13B" ~ "TYPE 9",
                            RMTYPE == "Oolitic" ~ "TYPE 10",
                            RMTYPE == "GB" ~ "TYPE 11",
                            TRUE ~ RMTYPE))

@@

S_RM_data <- S_RM_data %>%
  mutate_at("ID", str_replace, "VB", "") %>% # Remove "VB" from ID
  mutate(RMTYPE = case_when(
    RMTYPE == "Type 2" ~ "TYPE 2",
    RMTYPE == "Chalcedony" ~ "TYPE 1", # Altering first group names to final group names. Subtypes are maintained as before.
    RMTYPE == "TYPE 15" ~ "TYPE 3",
    RMTYPE == "TYPE 7" ~ "TYPE 4",
    RMTYPE == "TYPE 12" ~ "TYPE 5",
    RMTYPE == "TYPE 11" ~ "TYPE 6",
    RMTYPE == "TYPE 13" ~ "TYPE 7",
    RMTYPE == "TYPE 11F" ~ "TYPE 8",
    RMTYPE == "TYPE 13B" ~ "TYPE 9",
    RMTYPE == "Oolitic" ~ "TYPE 10",
    RMTYPE == "TYPE GB" ~ "TYPE 11",
    .default = RMTYPE))

@@

T_table <- phase_dataset %>%
  filter(RMTYPE !="INDET") %>% 
  mutate(RMTYPE = factor(RMTYPE, levels = c("TYPE 1","TYPE 2", "TYPE 3", "TYPE 4",
                                        "TYPE 5", "TYPE 6","TYPE 7", 
                                        "TYPE 8", "TYPE 9", "TYPE 10", "TYPE 11", "TL"))) %>%
  tabyl(RMTYPE, Level) %>%
  adorn_totals(c("row", "col")) %>%
  adorn_percentages("col") %>%
  adorn_pct_formatting(2) %>%
  adorn_ns("front")

@@

The three main used raw materials at the archaeological site of Vale Boi are quartz, greywacke, and chert. Other raw materials such as chalcedony, dolerite and schists have also been identified. Using the field database with all individually coordinated lithic artifacts, we compared the incidence of chert and chalcedony in opposition to the remaining lithics of other raw materials. @fig-rm-comparison shows that for the selected levels, chert represents only \~20-15% of the lithic assemblage. Based on previous lithic studies in the terrace area, the remaining 80% of other raw materials are mostly composed of quartz and greywacke. Although no systematic raw material research has been applied to the latter raw materials, they have been interpreted as local and readily available in the proximity of the site [@pereira_territory_2016-1]. Frequently, the quartz and greywacke assemblages include large amounts of shattered and/or unknapped chunks or slabs. Previous studies from the terrace area have also shown that when removing shatter and chunks, the percentages of chert become much more representative when compared to greywacke and quartz [@belmiro_gravettian-solutrean_2021; @belmiro_proto-solutrean_2020]. Based on this, despite the low amounts of individually plotted chert artifacts in the field database, chert continues to be one of the main raw materials to produce lithic stone tools, with complete knapping sequences and formal toolkits. The importance of chalcedony has also been shown in its use to produce formal tool kits during specific technocomplexes at the site [@belmiro_gravettian-solutrean_2021].

@@

context_short <- field_data %>% 
  filter(Level !="4B") %>%  # Filtering out levels with barely any artifacts
  mutate(Level = case_when(Level == "7C" ~ "7",
                           Level == 8 ~ "7", # Collapsing samples from Level 7C into 7 and two samples from Level 8 into 7, since 8 disappears in field_data
                           TRUE ~ Level)) %>% 
  mutate(Level = case_when(
    Level == 5 & Z > 24.1 ~ "4E/5",
    Level == 5 & Z < 24.1 ~ "5",
    TRUE ~ Level)) %>% 
  filter(!Level %in% c("3B", "5B", "5C"))

dataset_short <- phase_dataset %>% 
  select(ID, RMTYPE)
T_short <- full_join(context_short, dataset_short, by = "ID")
T_short["RMTYPE"][is.na(T_short["RMTYPE"])] <- "Other"

T_short <- T_short %>% 
 mutate(Rawmaterial = case_when(
    RMTYPE == "Other" ~ "Other",
    RMTYPE == "TYPE 1" ~ "Chalcedony",
    RMTYPE != "Other" | RMTYPE != "TYPE 1" ~ "Chert"))

@@

### Cherts congruent with local sources

Through the macroscopic and petrographic analysis, we identified 10 main chert/chalcedony groups, present throughout the stratigraphy in the Terrace and Shelter area (TABLE with groups). A single different group was identified in the Shelter area, with limited presence in the Terrace (n=2), and 14 varieties of chert and 2 chalcedony composed of one to three samples were identified in several levels of the Terrace and Shelter. The latter varieties were collapsed into a category named TL (Trace lithotypes). Through the comparison with the reference collections, we identified cherts congruent with the local sources and cherts which are not local.

TABLE WITH SHORT GROUP DESCRIPTION.

The raw materials which are congruent with local sources are the chalcedony and massive, micro-cryptocrystalline cherts (local_grid). The chalcedony (Type 1; A1) is a massive, fibrous chalcedony without allochems. It is characterized by a colorless and translucent appearance, often partially or completely white patinated. It is characterized by frequent large crystallizations, irregularities and fractures which impact the knapping. It is macroscopically and petrographically congruent with the hydrothermal chalcedony found in lenses in Triassic formations, also frequently characterized by irregularities and fractures. NEED chalcedony data from previous works from X.T.

The cherts are massive micro-cryptocrystalline quartzes from a marine environment. Macroscopically, these groups show variability which resulted in their separation in four different macroscopic groups (Group 2 to 5). In general, these are opaque to sub-translucent cherts, with colors varying brown, yellow, pale red and grey, with the frequent presence of spicules and iron oxides. The differences are in their colors, nodule morphology, frequency, and variety of bioclasts (Group 2 including triaxon sponge spicules and echinoderm spines), something already identified in the geological samples [@belmiro2023]. Petrographically, however, these cherts are homogeneous, except for a small number of artifacts with a banded variety within group 5. The macroscopic separation was maintained for its possible usefulness in identifying preferred types of nodules or macroscopic varieties through time and for future correlation with technological data. These cherts are characterized by wackestone textures, with accessory amounts of macrocrystalline quartz and fibrous chalcedony, frequently found replacing fossils and occasionally dolomite. Type 5 shows higher percentages of micrite/sparite. In all samples, fossils range from common to very frequent, albeit frequently unidentifiable. When identifiable in thin-section, they are sponge spicules (B2, C2, H2), with the rare occurrence of possible foraminifera (Type 2 and Type 5) and ostracod (H3) in a sample from type 5. Porosity varies between 10% and 1%, of vug type. Based on the previously described macroscopic and petrographic characteristics of the Lower Jurassic cherts of the Algarve [@belmiro2023], we interpret Groups 2, 3, 4 and 5 to be local, as they are congruent with the described characteristics. Groups 2 and 3 are congruent with the frequently available brown/yellow and red nodules and beds, ranging from uncommon to very frequent fossils. The macroscopic similarities of Groups 4 and 5 are more limited, as they show similarities to specific nodules, rarely found in the known outcrops, and mostly located in the Ferrel area. As such, we interpret these cherts as local, but from nodules and outcrops that are no longer frequently available or visible in the landscape. This is especially relevant, since the Ferrel outcrops are mostly destroyed and can be found in poor preservation conditions, and with small, broken nodules in the surface, frequently altered [@belmiro2023].

FIGURE: local_grid, variability of local chert macroscopic and petrographic characteristics

### Cherts not congruent with local sources

We identified several chert types with macroscopic and petrographic characteristics that are not congruent with the chert available in known outcrops in western Algarve (nonlocal_grid).

Group 6 (A1, B1, A2, B2) corresponds to a massive cryptocrystalline quartz with a mudstone texture, from a marine environment. This chert is translucent and shows a variety of colors, ranging from brown, red, and gray, which seem to vary in function of the artifact’s thickness. In unaltered samples, visible bioclasts are rare. The cortex is rounded, and rarely pebble-like, ranging from thin to medium thickness. Two sub-groups were individualized based on macroscopic characteristics such as color, translucency variation, and patterns, although they may simply reflect the use of specific nodules from within Type 6’s variability. Five thin sections were produced to better characterize the macroscopic variability of this group and sub-groups. The petrographic analysis shows the samples are characterized by a mudstone texture and massive microstructure. It is mainly composed of cryptocrystalline quartz (between 95-97% in the samples), with accessory fibrous chalcedony found replacing fossils or filling fractures, and in some samples macrocrystalline quartz, microcrystalline quartz, and micrite/sparite. Allochems are opaques, iron oxides, and bioclasts. Bioclasts are unidentifiable and poorly preserved, with rare identifiable fossils such as sponge spicules, replaced by fibrous chalcedony and rarely by microcrystalline quartz. Porosity is present between \<1-1%, of vug-type when identifiable.

FIGURE: nonlocal_grid, variability of non-local chert macroscopic and petrographic characteristics.
